AM205 Project - Traffic Flow Modelling\\ \Large{Microscopic Models of Traffic, Comparing the Intelligent Driver Model (with Adaptive Cruise Control) with the Human Driver Model}

### 2016-12-11 project_paper/paper.tex
updated images

\paragraph{}We then moved on to simulations of a platoon of cars with some cars being controlled by IDM-type models (in particular, Adaptive Cruise Control) and some cars being controlled by multi-car lookahead HDM models. After allowing the models to run for some time to develop their natural spacing, we then perturbed the acceleration of the lead car, causing it to brake and then resume again. We then observe both the magnitude and length of the volatility of the velocities in the following cars after the event. We find that in general, as one would expect, a larger proportion of IDM-controlled vehicles results in a quicker dissipation of the perturbation as well as a lowered probability of a vehicle collision.\\
\begin{figure}[H]
  \centering
  \fbox{\includegraphics[width=0.95\textwidth]{maximise_distance.png}}
  \caption{System of IDM cars on a circular track optimizing for maximum distance traveled. The more aggressive driving of the cars results in wave formation.}
\end{figure}

\section{Stability and Error Comparison of a Variety of Numerical Integration Schemes}
Because there is no analytical solution given for our highly-nonlinear problem, to analyze the error of our numerical integration schemes, we compared to the python \texttt{odeint} solver function. For large enough time steps, this shows us error decreasing on a log scale for each of our higher order methods. However, for small time steps, the `error` for 4th and 5th order methods converges, as the \texttt{odeint} solver presumably decides that 4th order is sufficiently accurate at these time steps and stops using higher order methods.\\
\begin{figure}
  \centering
  \fbox{\includegraphics[width=0.95\textwidth]{error.png}}
  \caption{Error to odeint decreases with accuracy of finite differencing methods.}
\end{figure}
\paragraph{}For our purposes, we desired to use higher order methods such that we could use larger timesteps and hopefully decrease running time of our programs while still maintaining stability in the differential equation systems. To display the effectiveness of these higher order methods, we ran one of our simple models for a variety of time steps over the same total time with the different solvers and noted the number of time steps at which each method stabilized. We notice here that we don't see an additional benefit from using RK5, which in this case we suspect is because the extremely complicated coefficients are contributing to existing numerical overflow issues.\\
\begin{table}[H]
\centering
\caption{Timesteps for Stability with T=500s}
\label{my-label}
\begin{tabular}{lc}
 Method &  Minimum Timesteps for Stability\\ \hline
 Heun's Method & 325 \\
 Runge-Kutta 3 &  275\\
 Runge-Kutta 4 &  250\\
 Runge-Kutta 5 & 250
\end{tabular}
\end{table}

@@

\section{Optimization of Traffic Systems}
\paragraph{}We set up a number of optimization programs to attempt to search for the optimal values of some of the many parameters for the system as a whole and for individual drivers for a number of cost functions, including maximizing displacement and maximizing comfort (minimizing jerk). The initial trials for these systems were run on IDM models, and the results are largely what one would suspect. When attempting to maximize total displacement, the cars drive quickly and close together, resulting in the creation of traffic waves in the system. When attempting to minimize jerk, the cars drive much more conservatively, as one would expect, never coming within the range of another car that would cause them to change acceleration. \\
\textbf{RESULTS HERE}
\paragraph{}We then attempted to optimize behavior for a single driver and were interested to see what implications this would have for the system as a whole. However, the following behaviors in the IDM are such that a single driver can tailgate and maximize target speed basically as much as we will allow without ever disrupting the rest of the system. Even when adding some stochasticity to the lead car, we were unable to see the rest of the system suffer from traffic waves due to the behavior of the second car, because that behavior would simply be quickly dissipated by the third car. In fact, it seems to be true generally of these systems that unless the aggressive driving behavior of a car causes other cars to collide behind it, the greater distance achieved by one car inevitably results in other cars also achieving a greater distance/average velocity. \\

@@

\begin{figure}[H]
  \centering
  \fbox{\includegraphics[width=0.95\textwidth]{optimize_s0.png}}
  \caption{Variation in Distance Traveled when Changing Safe Following Distance From Initial Value of 3.0}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \fbox{\includegraphics[width=0.95\textwidth]{optimize_t_change.png}}
  \caption{Variation in Distance Traveled when Changing Safe Following Time From Initial Value of 1.0}
\end{figure}

>>>>

\begin{figure}[H]
  \centering
  \fbox{\includegraphics[width=0.95\textwidth]{displacement_and_velocity_plot_model0.pdf}}
  \caption{System of 50 IDM-controlled cars on a circular track.}
\end{figure}

\begin{figure}[H]
  \centering
  \fbox{\includegraphics[width=0.95\textwidth]{displacement_and_velocity_plot_model2.pdf}}
  \caption{System of 50 ACC-controlled cars on a circular track.}
\end{figure}

\begin{figure}[H]
  \centering
  \fbox{\includegraphics[width=0.95\textwidth]{HDM_anticipate_1.pdf}}
  \caption{System of 50 HDM-controlled cars with one car-lookahead on a circular track.}
\end{figure}

\begin{figure}[H]
  \centering
  \fbox{\includegraphics[width=0.95\textwidth]{HDM_anticipate_5.pdf}}
  \caption{System of 50 HDM-controlled cars with five car-lookahead on a circular track. Note the significant decrease in volatility of velocity.}
\end{figure}

\paragraph{}We then moved on to simulations of a platoon of cars with some cars being controlled by IDM-type models (in particular, Adaptive Cruise Control) and some cars being controlled by multi-car lookahead HDM models. After allowing the models to run for some time to develop their natural spacing, we then perturbed the acceleration of the lead car, causing it to brake and then resume again. We then observe both the magnitude and length of the volatility of the velocities in the following cars after the event. We find that in general, as one would expect, a larger proportion of IDM-controlled vehicles results in a quicker dissipation of the perturbation as well as a lowered probability of a vehicle collision.\\

\begin{figure}[H]
  \centering
  \fbox{\includegraphics[width=0.95\textwidth]{blended_3_35_127pm.png}}
  \caption{System with 35 \% IDM cars on a straight road with acceleration of first car perturbed. System maintains a high degree of volatility in velocity long after the initial perturbation.}
\end{figure}

\begin{figure}[H]
  \centering
  \fbox{\includegraphics[width=0.95\textwidth]{blended_3_90_127pm.png}}
  \caption{System with 90 \% IDM cars on a straight road with acceleration of first car perturbed. System quickly dissipates the majority of the volatility in velocity.}
\end{figure}

\section{Stability and Error Comparison of a Variety of Numerical Integration Schemes}
Because there is no analytical solution given for our highly-nonlinear problem, to analyze the error of our numerical integration schemes, we compared to the python \texttt{odeint} solver function. For large enough time steps, this shows us error decreasing on a log scale for each of our higher order methods. However, for small time steps, the `error` for 4th and 5th order methods converges, as the \texttt{odeint} solver presumably decides that 4th order is sufficiently accurate at these time steps and stops using higher order methods.\\
\begin{figure}[H]
  \centering
  \fbox{\includegraphics[width=0.95\textwidth]{error.png}}
  \caption{Error to odeint decreases with accuracy of finite differencing methods.}
\end{figure}
\paragraph{}For our purposes, we desired to use higher order methods such that we could use larger timesteps and hopefully decrease running time of our programs while still maintaining stability in the differential equation systems. To display the effectiveness of these higher order methods, we ran one of our simple models for a variety of time steps over the same total time with the different solvers and noted the number of time steps at which each method stabilized. We notice here that we don't see an additional benefit from using RK5, which in this case we suspect is because the extremely complicated coefficients are contributing to existing numerical overflow issues.\\
\begin{table}[H]
\centering
\caption{Timesteps for Stability with T=500s}
\label{my-label}
\begin{tabular}{lc}
 Method &  Minimum Timesteps for Stability\\ \hline
 Heun's Method & 325 \\
 Runge-Kutta 3 &  275\\
 Runge-Kutta 4 &  250\\
 Runge-Kutta 5 & 250
\end{tabular}
\end{table}

@@

\section{Optimization of Traffic Systems}
\paragraph{}We set up a number of optimization programs to attempt to search for the optimal values of some of the many parameters for the system as a whole and for individual drivers for a number of cost functions, including maximizing displacement and maximizing comfort (minimizing jerk). The initial trials for these systems were run on IDM models, and the results are largely what one would suspect. When attempting to maximize total displacement, the cars drive quickly and close together, resulting in the creation of traffic waves in the system. When attempting to minimize jerk, the cars drive much more conservatively, as one would expect, never coming within the range of another car that would cause them to change acceleration. \\

\begin{figure}[H]
  \centering
  \fbox{\includegraphics[width=0.95\textwidth]{maximise_distance.png}}
  \caption{System of IDM cars on a circular track optimizing for maximum distance traveled. The more aggressive driving of the cars results in wave formation.}
\end{figure}

\paragraph{}We then attempted to optimize behavior for a single driver and were interested to see what implications this would have for the system as a whole. However, the following behaviors in the IDM are such that a single driver can tailgate and maximize target speed basically as much as we will allow without ever disrupting the rest of the system. Even when adding some stochasticity to the lead car, we were unable to see the rest of the system suffer from traffic waves due to the behavior of the second car, because that behavior would simply be quickly dissipated by the third car. In fact, it seems to be true generally of these systems that unless the aggressive driving behavior of a car causes other cars to collide behind it, the greater distance achieved by one car inevitably results in other cars also achieving a greater distance/average velocity. \\

@@

\begin{figure}[H]
  \centering
  \fbox{\includegraphics[width=0.95\textwidth]{optimize_s0.png}}
  \caption{Variation in Distance Traveled when Changing Safe Following Distance Relative to Initial Value of 3.0}
\end{figure}

@@

\begin{figure}[H]
  \centering
  \fbox{\includegraphics[width=0.95\textwidth]{optimize_t_change.png}}
  \caption{Variation in Distance Traveled when Changing Safe Following Time Relative to Initial Value of 1.0}
\end{figure}
